\section{The Audacious Proposal}

Sentence three: \emph{``We propose a new simple network architecture, the Transformer, based solely on attention mechanisms, dispensing with recurrence and convolutions entirely.''}

Read that sentence again.
Every word is chosen carefully.

\textbf{Simple.}
They are not claiming to have built something more complicated.
They are claiming the opposite: that their architecture is \emph{simpler} than what came before.
In a field where progress usually means adding more moving parts, they removed parts.

\textbf{Solely.}
Not ``primarily.''
Not ``mostly.''
Solely.
The Transformer uses attention and nothing else as its core mechanism.

\textbf{Dispensing with recurrence and convolutions entirely.}
This is the mic drop.
Recurrence --- the defining feature of RNNs, the sequential one-word-at-a-time processing that the entire field was built on --- gone.
Convolutions --- the local-window approach that was the main alternative --- also gone.

To appreciate how bold this was, you have to understand that in 2017, the idea of processing sequences \emph{without} recurrence was a little like proposing that you could drive a car without wheels.
Recurrence was how you handled sequential data.
That was not a design choice; it was, as far as most people were concerned, a fact of nature.

And yet the authors are saying: actually, no.
If you have a powerful enough attention mechanism --- one that lets every element in the sequence look at every other element simultaneously --- you do not need to process things in order at all.
You can look at the whole sequence at once.

The name they gave it --- Transformer --- has become so ubiquitous that it is easy to forget it was a new coinage.
GPT stands for Generative Pre-trained \emph{Transformer}.
BERT stands for Bidirectional Encoder Representations from \emph{Transformers}.
Every major language model since 2018 is a Transformer or a direct descendant of one.

This sentence is where the paper plants its flag.
The next three sentences are about proving that the flag was planted in the right place.
