\section{2017: Attention Is All You Need}

\subsection{The Paper That Changed Everything}

In June 2017, a team of eight researchers at Google published a paper with a deliberately provocative title: \emph{``Attention Is All You Need''} \cite{vaswani2017attention}.

The paper introduced the \textbf{Transformer} architecture.
It dispensed entirely with recurrence --- no RNNs, no LSTMs, no processing one token at a time.
Instead, it proposed that \emph{attention} was sufficient to model language.

The results were immediate and dramatic: better performance on translation benchmarks, faster training, and a model that scaled gracefully with more data and more computation.

\subsection{What is Attention?}

Before Transformers, researchers had already been experimenting with attention as an add-on to RNNs \cite{bahdanau2014neural}.
The idea: when translating a word, instead of relying on a single compressed hidden state, let the model look back at all the input words and decide which ones are most relevant right now.

The Transformer took this idea and made it the entire foundation of the model.

\begin{tcolorbox}[colback=blue!5!white, colframe=blue!40!black, title=Attention in Plain English]
Imagine you are translating the French sentence ``Le chat noir dort'' into English.
When you are about to write the word ``black,'' your brain naturally focuses on ``noir'' --- not the whole sentence equally.

Attention does the same thing mathematically.
For each word being produced, the model computes a \emph{score} between that word and every word in the input, then uses those scores to create a weighted combination.
High-scoring words contribute more; low-scoring words contribute less.
\end{tcolorbox}

\subsection{Self-Attention: The Key Innovation}

The Transformer introduced \textbf{self-attention}: the model attends not just to an input sequence, but to its own sequence --- every word attends to every other word simultaneously.

This means the model can capture long-range dependencies in a single step.
The word ``it'' at the end of a long sentence can directly attend to the noun it refers to at the beginning, without the signal having to travel through dozens of sequential steps.

\begin{tcolorbox}[colback=green!5!white, colframe=green!40!black, title=Why Self-Attention Matters]
Consider: ``The trophy didn't fit in the suitcase because \textbf{it} was too big.''

What does ``it'' refer to? The trophy or the suitcase?
A human reader knows it's the trophy --- because if the suitcase were too big, it would have fit.

Self-attention lets the model directly compare ``it'' with every other word in the sentence and learn, from training data, that ``too big'' points back to the trophy.
An RNN has to carry this inference through sequential hidden states; self-attention resolves it in a single computation.
\end{tcolorbox}

\begin{figure}[htb]
\centering
\begin{tikzpicture}[>=stealth]
  \def\sep{1.55}
  % Word boxes
  \foreach \i/\w in {0/The, 1/trophy, 2/{didn't}, 3/fit, 4/because, 5/it, 6/was, 7/too, 8/big} {
    \node[rectangle, draw=black!70, fill=blue!8, rounded corners=2pt,
          minimum width=1.35cm, minimum height=0.55cm, font=\footnotesize]
      (w\i) at (\i*\sep, 0) {\w};
  }
  % Weak attention from "it" (w5) to other words
  \foreach \i in {0,2,4} {
    \draw[->, line width=0.5pt, gray!35] (w5.north) to[bend right=45] (w\i.north);
  }
  \foreach \i in {6,7,8} {
    \draw[->, line width=0.5pt, gray!35] (w5.north) to[bend left=45] (w\i.north);
  }
  % Medium attention to "fit" (w3)
  \draw[->, line width=2pt, blue!55] (w5.north) to[bend right=45] (w3.north);
  % Strong attention to "trophy" (w1)
  \draw[->, line width=4pt, blue!80] (w5.north) to[bend right=45] (w1.north);
  % Legend
  \node[font=\footnotesize\itshape, text=gray, align=center]
    at (4*\sep, -1.1) {Arrow thickness $\propto$ attention weight from ``\textbf{it}''};
\end{tikzpicture}
\caption{Self-attention resolving a pronoun. When processing ``it,'' the model computes an attention score over every other word. Thicker arrows indicate stronger attention. The model learns to strongly attend to ``trophy'' --- the correct referent --- with no explicit grammatical rules required.}
\end{figure}

\subsection{Parallelism: Training at Scale}

The other revolution was practical.
Because self-attention processes all positions simultaneously, Transformers are \emph{massively parallelizable}.
Every GPU core can work on a different part of the computation at the same time.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[>=stealth, thick]
  % ---- LEFT: RNN sequential ----
  \node[font=\small\bfseries] at (1.65, 1.8) {RNN (sequential)};
  \node[rectangle, draw=red!60, fill=red!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (r1) at (0,   0) {$w_1$};
  \node[rectangle, draw=red!60, fill=red!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (r2) at (1.1, 0) {$w_2$};
  \node[rectangle, draw=red!60, fill=red!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (r3) at (2.2, 0) {$w_3$};
  \node[rectangle, draw=red!60, fill=red!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (r4) at (3.3, 0) {$w_4$};
  \draw[->, red!70] (r1) -- (r2);
  \draw[->, red!70] (r2) -- (r3);
  \draw[->, red!70] (r3) -- (r4);
  \node[font=\footnotesize, text=red!70, align=center] at (1.65, -1.0)
    {step 1 $\to$ step 2 $\to$ step 3 $\to$ step 4};

  % ---- DIVIDER ----
  \draw[dashed, gray!40, thin] (4.05, -1.6) -- (4.05, 2.1);

  % ---- RIGHT: Transformer parallel ----
  \node[font=\small\bfseries] at (6.3, 1.8) {Transformer (parallel)};
  \node[rectangle, draw=green!60!black, fill=green!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (t1) at (4.6, 0.85) {$w_1$};
  \node[rectangle, draw=green!60!black, fill=green!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (t2) at (5.5, 0.85) {$w_2$};
  \node[rectangle, draw=green!60!black, fill=green!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (t3) at (6.4, 0.85) {$w_3$};
  \node[rectangle, draw=green!60!black, fill=green!8, rounded corners=2pt,
        minimum width=0.85cm, minimum height=0.55cm, font=\footnotesize] (t4) at (7.3, 0.85) {$w_4$};
  \node[rectangle, draw=green!60!black, fill=green!18, rounded corners=3pt,
        minimum width=3.6cm, minimum height=0.7cm, font=\small]
    (att) at (5.95, -0.2) {Attention};
  \draw[->, green!60!black] (t1.south) -- (att.north -| t1);
  \draw[->, green!60!black] (t2.south) -- (att.north -| t2);
  \draw[->, green!60!black] (t3.south) -- (att.north -| t3);
  \draw[->, green!60!black] (t4.south) -- (att.north -| t4);
  \node[font=\footnotesize, text=green!50!black, align=center] at (5.95, -1.0)
    {all 4 words at once, 1 step};
\end{tikzpicture}
\caption{Sequential vs.\ parallel processing. An RNN must process tokens one at a time, taking $n$ steps for a sequence of length $n$. A Transformer's self-attention computes relationships between all tokens simultaneously in a single step, enabling training on far larger datasets.}
\end{figure}

\begin{tcolorbox}[colback=gray!5!white, colframe=gray!50!black,
  title={Python analogy: the difference in two programs}]
\textbf{RNN --- sequential. Step 2 is blocked until step 1 finishes:}\\[4pt]
{\small\ttfamily
h = initial\_state\\
for word in ["The", "cat", "sat", "down"]:\\
\hspace{2em}h = rnn\_step(h, word)\quad\# h changes each iteration;\\
\hspace{2em}\# the next iteration \emph{must} wait for this one to complete
}

\medskip\rule{\linewidth}{0.4pt}\medskip

\textbf{Transformer --- parallel. All words launch at the same time:}\\[4pt]
{\small\ttfamily
words = ["The", "cat", "sat", "down"]\\
output = attention(words)\quad\# no loop, no waiting;\\
\# every word is processed in one simultaneous operation
}

\medskip
{\small The RNN loop \emph{cannot} be parallelized --- each iteration depends on the previous value of \texttt{h}. The Transformer call \emph{can} be split across thousands of GPU cores at once, because no word's computation depends on another's result.}
\end{tcolorbox}

This meant that for the first time, language models could be trained on enormous datasets in reasonable time.
Scale was no longer the enemy --- it became the strategy.

Researchers quickly discovered an empirical pattern: \emph{bigger models trained on more data consistently performed better}.
This observation, later formalized as \textbf{scaling laws} \cite{kaplan2020scaling}, would drive the next phase of the field.
