\section{Deep Learning Is Not Dead}

\subsection{A Common Misconception}

When ChatGPT became a cultural phenomenon, a narrative took hold in tech circles: \emph{the era of deep learning is over; we are now in the era of large language models.}

This is a category error.

Large language models \emph{are} deep learning.
A Transformer is a deep neural network.
It is trained by gradient descent with backpropagation.
It learns by adjusting millions or billions of numerical weights based on prediction errors.
Every concept introduced in this paper --- neurons, layers, weights, activation functions, backpropagation --- applies directly to GPT-4, Claude, and DeepSeek.

LLMs are not a departure from deep learning.
They are its current frontier.

\subsection{The Larger Ecosystem}

Deep learning is also very much alive outside of language.

\begin{itemize}
    \item \textbf{Computer vision} continues to advance, with models like Vision Transformers (ViT) \cite{dosovitskiy2020image} applying the Transformer architecture directly to image patches.
    \item \textbf{Protein folding}: AlphaFold 2 \cite{jumper2021highly}, released by DeepMind in 2020, used deep learning to predict the 3D structure of proteins from their amino acid sequences --- solving a fifty-year-old biology problem.
    \item \textbf{Generative image models}: Diffusion models, which power tools like Stable Diffusion and DALL-E, are deep neural networks trained to iteratively denoise images.
    \item \textbf{Scientific computing}: Deep learning is being applied to weather forecasting, drug discovery, materials science, and particle physics.
    \item \textbf{Robotics}: Neural networks now control robot arms that can assemble objects, fold laundry, and navigate unstructured environments.
\end{itemize}

The field has not narrowed --- it has expanded.

\subsection{Why the Misconception Persists}

Part of the confusion stems from how the public encounters AI.
For most people, AI \emph{is} the chatbot.
The underlying architecture is invisible.

But the researchers building these systems know exactly where they are standing.
When OpenAI trains a new model, they are doing deep learning.
When DeepSeek builds a Mixture of Experts architecture, they are doing deep learning.
When Anthropic fine-tunes Claude with constitutional principles, they are doing deep learning.

\begin{tcolorbox}[colback=green!5!white, colframe=green!40!black, title=A Useful Analogy]
Saying ``deep learning is dead because of LLMs'' is like saying ``electricity is dead because of smartphones.''

Smartphones run on electricity.
LLMs run on deep learning.

The underlying foundation did not get replaced --- it got more powerful, more refined, and more widely deployed.
\end{tcolorbox}

\subsection{What Actually Changed}

What changed after 2017 was not the paradigm --- it was the \emph{scale and the ambition}.

Before Transformers, practitioners asked: \emph{What specific task can we train a model for?}
After Transformers and scaling laws, researchers started asking: \emph{What happens if we train a single model on essentially everything?}

That shift in question is what produced the LLMs we see today.
And it happened entirely within the framework of deep learning.
